Graphical Tests for Power Comparison of Competing Designs

### 2012-03-31 setup.tex
tweaked language slightly

Two recent data analyses motivated the setup of this experiment. The first one arises from an analysis of US flight traffic data, and a finding related to how wind direction affects the efficiency of an airport.  The second one arose during the review of a paper claiming that it was impossible to statistically test for differences in center between two samples where sample size was small and from a non-normal population, with application to studying toxic waste sites. Using lineups containing dotplots we found it was possible to detect a difference, and we were curious to see if other display types (density, histogram, boxplot) might be better than dotplots.

@@

 Each participant is shown ten lineups and asked to identify the plot that is the most different from the other plots. In the second study the `difference' is further specified as `In which plot is the blue group the furthest to the right?". 
 In both studies, though, the question is placed prominently above each lineup as a reminder of what participants are asked to look for. 
  Additionally, participants are asked for a reason for their choice. They can select from a list of choices particular to the lineup design, why they chose this particular plot as well as how confident they are on a scale of one to five. 
 Personal information such as age group, gender, and education is collected on a voluntary basis. 
The amount of time it takes the individual to answer is also recorded. 
This  allows us to assess how design choices affect decoding \citet{cleveland:1994} in terms of accuracy and speed.

 In order to avoid people `gaming the system' as found in earlier studies \cite{heer:2010, kosara:2010}, participants were shown specially prepared lineups as a baseline of performance -- if participants fail to answer some of these baseline lineups, we require at least two correct answers for the remaining plots, which is indicative of a better-than-guessing performance, as inclusion criterion for the data evaluation.

@@

The data contains all flights \cite{rita} in and out of Seattle/Tacoma International Airport (SEA) between July 2008 and June 2011. As a measure of airport efficiency we are using time between successive wheel-events (time at wheel take-off or touch-down), which we found to be independent of airline carrier and operating hours, as long as we restricted the data to `regular' operating hours between 7 am and midnight and `regular' weather conditions -- i.e. we eliminated records associated with the top 5\% wind speed measurements \cite{noaa-weather}, leaving us with just under 0.5 Million flights.

In this scenario, we are dealing with the null hypothesis $H_0: $ wind direction has no effect on efficiency and the alternative $H_a:$ wind direction does affect efficiency.

@@

In spite of the contextual circular property of wind direction, the pattern in efficiency did not seem to stand out well during the exploration process, which led us to using a corresponding  Euclidean design. Both designs were additionally equipped with a reference line at 50\%. 
Figure \ref{layouts} shows an overview of all four chart types in the study. During the exploration of the data, it became clear that Seattle airport functions at it best with winds coming from the East to Southeast, while winds from the West seemed to be most troublesome, resulting in a wave-like pattern in the Euclidean charts and a shift in center for the polar charts. Since runways can be used in both directions, changing the runway usage according to dominant wind direction for that day or time of the year might be a feasible solution in gaining efficiency.

An additional factor we were interested in this particular situation was to assess, how much of the data we actually needed to use in a design to have observers pick out the pattern. Clearly, a design is more efficient, if a smaller sample size is sufficient for showing the presence of a relationship. In order to investigate this, we took different size samples of the data and created lineup plots of all four designs for these subsets. The effect of sampling size on the displays mostly stems from the additional variability introduced when using small samples, which might hide the pattern, if it is not displayed prominently.

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=\linewidth]{euclid-offset} 
   \includegraphics[width=\linewidth]{polar-offset} 
   \caption{ \label{fig:offset} (Top) Euclidean charts of the same sample, with shifts in wind direction by 90, 180, and 270 degrees. The shift by 90 and 270 leads to a `valley' versus a `mountain' pattern, whereas a shift by 180 degrees inverts the wave pattern from `down-up' to `up-down'.\hfill\newline
      (Bottom) Shifts in wind direction in the polar chart lead to rotations of the display by 90 degrees.}
\end{figure}

 This results in a shift of the wave in Euclidean charts and a rotation in the polar charts as can be seen in figure \ref{fig:offset}. Our initial hypothesis was that this would have no effect on the polar charts, but might have a deteriorating effect on the Euclidean charts, in the case that the peak or the valley of the wave was at either end of the $x$ axis, i.e. an offset of 90 or 270 degrees. 
While these simple perturbations and different sample sizes allow us to get insight into different aspects of the designs, they also allow us to get multiple responses from each observer to assess individual ability without the need to go outside the framework of the original data, i.e. we leave the joint relationship between $x$ and $y$ essentially unchanged. For all of these combinations we produced two replicates, resulting in a total of 192 different lineups ($ 4 \text{ designs} \times 6 \text{ sample sizes} \times 4 \text{ offsets} \times 2 \text{ reps})$.
Table \ref{tbl:treatment} gives an overview of the number of times lineups in each combination of design and sample size were shown, and in how many of them the data plot was correctly identified.

@@

In a second experiment, we conducted a simulation study to investigate the power of four competing designs, as shown in figure \ref{fig:expii}, in assessing a mean shift between distributions. The null hypothesis of each lineup is $H_o:$ centers of the two groups are the same, and the alternative hypothesis is $H_a:$ center of the blue group is shifted to the right.

As covariates for this experiment, we considered the {\it size of the shift}, $d$, between the means of the distributions, with $d=0.4, 0.6, 0.8. 1.0,$ and $1.2$, as well as the sample size.  Since we are dealing with two groups of points, we varied both the size of the larger group, $n_1 = 15, 45$, and 135, as well as the relative size of the second group, with $n_2 = 1/3, 2/3$ and $3/3$ of $n_1$. 
For each of these combinations, we created three replicate data sets with values from an Exponential distribution with rate parameter $\lambda=1$ (for the first group) and rate parameter $\lambda_2=1/(d+1)$, which corresponds to a mean shift of $d$, for the second groups. 
Based on each data set we created four lineup charts, one for each of the competing designs. This resulted in 540 different lineup charts overall, which we sorted by $p$-value corresponding to the difference in means between the two groups in the simulated data. According to the $p$-value, we associated a `difficulty' level with lineups from 1 to 9 (easy to hard). Ten lineup plots were shown to each participant - one out of each of the difficulty levels and one additional reference chart with a particularly easy lineup  to allow for data quality checks.
\begin{figure} [hbtp]
   \centering
   \includegraphics[width=0.45\linewidth]{hist-conc} 
   \includegraphics[width=0.45 \linewidth]{boxplot-conc} \\
   \includegraphics[width=0.45 \linewidth]{density-conc} 
   \includegraphics[width=0.45 \linewidth]{dotplot-conc} 
   \caption{Overview of all four competing designs for displaying differences in distributions.}
   \label{fig:expii}
\end{figure}

>>>>

Two recent data analyses motivated the setup of this experiment. The first one arises from an analysis of US flight traffic data, and a finding related to how wind direction affects the efficiency of an airport.  The second one arose during the review of a paper claiming that it was impossible to statistically test for differences in center between two samples where sample size was small and from a non-normal population, with application to studying toxic waste sites. Using lineups containing dotplots we found it was possible to detect a difference, and we were curious to see if other display types (density, histogram, boxplot) might compare with dotplots.

@@

Each participant is shown ten lineups and asked to identify the plot that is the most different from the other plots. In the second study the `difference' is further specified as `In which plot is the blue group the furthest to the right?". 
 In both studies, though, the question is placed prominently above each lineup as a reminder of what participants are asked to look for.  Additionally, participants are asked for a reason for their choice. They can select from a list of choices particular to the lineup design, why they chose this particular plot as well as how confident they are on a scale of one to five. Personal information on age group, gender, and education is collected on a voluntary basis.  The amount of time it takes the individual to answer is also recorded.  This  allows us to assess how design choices affect decoding \citet{cleveland:1994} in terms of accuracy and speed.

 In order to avoid people `gaming the system' as found in earlier studies \cite{heer:2010, kosara:2010}, two of the ten lineups participants were shown were specially prepared as a baseline of performance -- if participants fail to answer some of these baseline lineups, we require at least two correct answers for the remaining plots, which is indicative of a better-than-guessing performance, as inclusion criterion for the data evaluation.

@@

The data contains all flights \cite{rita} in and out of Seattle/Tacoma International Airport (SEA) between July 2008 and June 2011. As a measure of airport efficiency we are using time between successive wheel-events (time at wheel take-off or touch-down), which we found to be independent of airline carrier and operating hours, as long as we restricted the data to `regular' operating hours between 7 am and midnight and `regular' weather conditions -- i.e. we eliminated records associated with the top 5\% wind speed measurements \cite{noaa-weather}, leaving us with just under half a million flights.

In this scenario, all of the lineups are testing the null hypothesis

\vspace{-0.1in}
\begin{quote}
$H_0: $ {\em wind direction has no effect on efficiency }
\end{quote}
\vspace{-0.1in}

\noindent against the alternative $H_a:$ wind direction does affect efficiency.

@@

In spite of the contextual circular property of wind direction, the pattern in efficiency did not seem to stand out well during the exploration process, which led us to using a corresponding  Euclidean design. Both designs were additionally equipped with a reference line at 50\%. 
Figure \ref{layouts} shows an overview of all four chart types in the study. During the exploration of the data, it became clear that Seattle airport functions at it best with winds coming from the east to southeast, while winds from the west seemed to be most troublesome, resulting in a wave-like pattern in the Euclidean charts and a shift in center for the polar charts. Since runways can be used in both directions, changing the runway usage according to dominant wind direction for that day or time of the year might be a feasible solution in gaining efficiency.

An additional factor we were interested in this particular situation was to assess, how much of the data we actually needed to use in a design to have observers pick out the pattern. Clearly, a design is more efficient, if a smaller sample size is sufficient for showing the presence of a relationship. In order to investigate this, we took different size samples of the data and created lineup plots of all four designs for these subsets. The effect of sample size on the displays mostly stems from the additional variability introduced when using small samples, which might hide the pattern, if it is not displayed prominently.

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=\linewidth]{euclid-offset} 
   \includegraphics[width=\linewidth]{polar-offset} 
   \caption{ \label{fig:offset} (Top) Euclidean charts of the same sample, with shifts in wind direction by 90, 180, and 270 degrees. The shift by 90 and 270 leads to a `valley' versus a `mountain' pattern, whereas a shift by 180 degrees inverts the wave pattern from `down-up' to `up-down'. %\hfill\newline
      (Bottom) Shifts in wind direction in the polar chart lead to rotations of the display by 90 degrees.}
\end{figure}

 This results in a shift of the wave in Euclidean charts and a rotation in the polar charts as can be seen in Figure \ref{fig:offset}. Our initial thinking was that this would have no effect on the polar charts, but might have a deteriorating effect on the Euclidean charts, in the case that the peak or the valley of the wave was at either end of the $x$ axis, i.e. an offset of 90 or 270 degrees. 
While these simple perturbations and different sample sizes allow us to get insight into different aspects of the designs, they also allow us to get multiple responses from each observer to assess individual ability without the need to go outside the framework of the original data, i.e. we leave the joint relationship between $x$ and $y$ essentially unchanged. For all of these combinations we produced two replicates, resulting in a total of 192 different lineups ($ 4 \text{ designs} \times 6 \text{ sample sizes} \times 4 \text{ offsets} \times 2 \text{ reps})$.
Table \ref{tbl:treatment} gives an overview of the number of times lineups in each combination of design and sample size were shown, and in how many of them the data plot was correctly identified.

@@

In a second experiment, we conducted a simulation study to investigate the power of four competing designs, as shown in Figure \ref{fig:expii}, in assessing a mean shift between distributions. The null hypothesis of each lineup is

\vspace{-0.1in}
\begin{quote}
$H_o:$ centers of the two groups are the same,
\end{quote}
\vspace{-0.1in}

\noindent and the alternative hypothesis is $H_a:$ center of the blue group is shifted to the right.

As covariates for this experiment, we considered the {\it size of the shift}, $d$, between the means of the distributions, with $d=0.4, 0.6, 0.8. 1.0,$ and $1.2$, as well as the sample size.  Since we are dealing with two groups of points, we varied both the size of the larger group, $n_1 = 15, 45$, and 135, as well as the relative size of the second group, with $n_2 = 1/3, 2/3$ and $3/3$ of $n_1$. 
For each of these combinations, we created three replicate data sets with values from an exponential distribution with rate parameter $\lambda=1$ (for the first group) and rate parameter $\lambda_2=1/(d+1)$, which corresponds to a mean shift of $d$, for the second groups. 
Based on each data set we created four lineup charts, one for each of the competing designs. This resulted in 540 different lineup charts overall, which we sorted by $p$-value corresponding to the difference in means between the two groups in the simulated data. According to the $p$-value, we associated a `difficulty' level with lineups from 1 to 9 (easy to hard). Ten lineup plots were shown to each participant - one out of each of the difficulty levels and one additional reference chart with a particularly easy lineup  to allow for data quality checks.
\begin{figure} [hbtp]
   \centering
   \includegraphics[width=0.45\linewidth]{hist-conc} 
   \includegraphics[width=0.45 \linewidth]{boxplot-conc} \\
   \includegraphics[width=0.45 \linewidth]{density-conc} 
   \includegraphics[width=0.45 \linewidth]{dotplot-conc} 
   \caption{Overview of all four competing designs for displaying differences in distributions.}
   \label{fig:expii}
\end{figure}
